---
title: "Assignment #5, Initial Data Analysis"
author: "YOUR NAME"
date: 'Updated on `r format(Sys.Date(), "%B %d, %Y")`'
output: 
   html_document:
     theme: journal
     highlight: textmate
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
```

### Load Libraries

```{r libraries, warning = FALSE, message = FALSE}
## Load libraries
library(tidyverse)
library(sf)
library(tmap)
library(psych)
library(knitr)
library(kableExtra)
```

### Read Data

```{r readdata, warning = FALSE, message = FALSE}
# Read NC Zip Code spatial data
nc_zip_sp <- read_sf("../Data/NC_ZipCode_Polygons.gpkg")

# Read NC Zip Code COVID data
nc_zip_covid <- read_csv("../Data/NC_ZipCode_Covid.csv")
```

### Data Wrangling

```{r wrangle, warning = FALSE, message = FALSE}
####
#### Table join the attributes of nc_zip_covid to nc_zip_sp
#### 
#### Write output to a new object called   nc_zip_covid_sp
####

# Your code here


####
#### Calculate new columns with Case and Death Rates per 1,000 people
#### 
#### Use CASERATE and DEATHRATE for the new column names
####
#### Formula is, for example,  1000 * cases / people
####

# Your code here


####
#### Calculate the statewide Case and Death Rates per 1,000 people
#### 

#### First, use summarize() to sum the cases, deaths, and population
####
#### USE THE SAME FIELD NAMES IN THE SUMMARIZED OUTPUT
####
nc_covid <- nc_zip_covid_sp |>
  st_drop_geometry() |>    # This function removes spatial column
  summarize()  ### You are required to finish this function!
  

####
#### Next, Calculate new columns with Case and Death Rates per 1,000 people
#### 
#### Use CASERATE and DEATHRATE for the new column names
####
#### Formula is, for example,  1000 * cases / people
####

# Your code here

```

### Descriptive Statistics

```{r summarize, warning = FALSE, message = FALSE}
# Create summary information table
summary_table <- nc_zip_covid_sp |> 
  st_drop_geometry() |>                   # Remove spatial column
  select(-ZIP_CODE, -PO_NAME, -POP) |>    # Remove some columns
  na.omit() |>                            # Remove obs with NA
  describe(fast = TRUE)                   # Create summary table

# Remove vars, std error from summary table
summary_table <- summary_table |>
  select(-vars, -se)

## Print nice version of the table
kable(summary_table, 
      digits = 1,
      format.args = list(big.mark = ",",
                         scientific = FALSE,
                         drop0trailing = TRUE),
      caption = "Summary of North Carolina COVID-19 Cumulative Cases, Deaths, Case Rates (per 1,000 residents), and Death Rates (per 1,000 residents) by Zip Code") %>% 
  kable_styling(bootstrap_options = c("striped", 
                                      "hover", 
                                      "condensed", 
                                      "responsive"), 
                full_width = F)


```


### Histogram 1

```{r histogram1, warning = FALSE, message = FALSE, fig.width = 7, fig.height = 4}
## plot histogram
ggplot(nc_zip_covid_sp, 
       aes(x = CASERATE)) + 
  geom_histogram(binwidth = 0.2) +
  labs(x = "Cumulative COVID-19 cases per 100,000 People",
       y = "Count",
       title = "Histogram of COVID-19 Case Rates") +
  theme_minimal()

```


### Histogram 2

```{r histogram2, warning = FALSE, message = FALSE, fig.width = 7, fig.height = 4}
## Copy/paste code from Histogram 1 above but change 
## the binwidth to improve the histogram
## plot histogram


```

### Histogram 3

```{r histogram3, warning = FALSE, message = FALSE, fig.width = 7, fig.height = 4}
## plot histogram
ggplot(nc_zip_covid_sp, 
       aes(x = CASES)) + 
  geom_histogram(binwidth = 50) +
  xlab("Cumulative COVID-19 cases") +
  ggtitle("Histogram of COVID-19 Cases") +
  theme_minimal()

```

### Choropleth Map 1

```{r map1, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 5}
## Fill in the missing blanks below to make a map of the 
## COVID Case Rates per 1000 people. Use jenks for the 
## choropleth classification and YlOrRd for the color
## palette

## Plot map
tm_shape(______) + 
  tm_polygons(______, 
              style = ______, 
              palette = ______,
              lwd = 0.25,
              border.col = "black",
              border.alpha = 0.5,
              title = "Cumulative COVID-19 Case Rate (per 1,000 people)")
```


### Choropleth Map 2

```{r map2, warning = FALSE, message = FALSE, fig.width = 9, fig.height = 5}
## Copy/paste code from Choropleth Map 1 (above), but change
## the choropleth classification to quantile
## Plot map


```


******

##### Question 1. Write one sentence that **uses information from the table above** to describe the central tendency and dispersion of COVID-19 Case Rates (per 1,000 people) by Zip Code. (1 sentence, using at least one measure of central tendency and one measure of dispersion)

YOUR ANSWER HERE

  
******


##### Question 2. Describe the differences between the data distributions found in Histogram 2 and Histogram 3? (1-2 sentences)

YOUR ANSWER HERE

  
******


##### Question 3. Why is mapping COVID-19 Case Rates more appropriate than mapping the raw number of COVID-19 Cases? (2-3 sentences) 

YOUR ANSWER HERE

  
******


##### Question 4. Note that Choropleth Map 1 and 2 map the exact same data but use different choropleth classification schemes. In your opinion, which is easier for you to interpret and why is it easier? (2-3 sentences) 

YOUR ANSWER HERE

  
******